---
title: "Exploratory Analysis of LA City Employee Payroll"
author: "Andrew Winter"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


I am looking at [open data](https://data.lacity.org/) from the city of Los Angeles. [This data set](https://controllerdata.lacity.org/Payroll/City-Employee-Payroll/pazn-qyym) hasn't been updated since June 2019, but it's sufficient for EDA sample code. I used their Socrata API to access the data.


```{r packages, message=FALSE}
# Load required packages
library(dplyr)
library(purrr)
library(stringr)
library(ggplot2)
library(httr)
library(scales)
```


```{r basic procedure, message=FALSE}
payroll_base <- "https://controllerdata.lacity.org/resource/pazn-qyym.csv"
payroll_base_response <- httr::GET(
                           payroll_base,
                           add_headers("X-App-Token" =
                             Sys.getenv("SOCRATA_APP_TOKEN"))
                           )
content(payroll_base_response, as = "parsed")
```


By default, only 1,000 rows are returned. I can take the API endpoint and add limit, offset, and order parameters.


```{r parameters example}
# For example, three rows only
payroll_base %>%
  paste0("?$limit=3&$offset=0&$order=row_id") %>%
  GET(add_headers("X-App-Token" = Sys.getenv("SOCRATA_APP_TOKEN")))
```


```{r set of functions 1}
# Functions to reduce copy and paste code
limit_path <- function(lim = 50, off = 0) {
  paste0(payroll_base, "?$limit=", lim, "&$offset=", off, "&$order=row_id")
}
get_parse <- function(url) {
  GET(url, add_headers("X-App-Token" = Sys.getenv("SOCRATA_APP_TOKEN"))) %>%
    content(as = "parsed")
}
```

```{r list and df, warning=FALSE, message=FALSE}
# Believe I can only get 50,000 rows at a time
# First batch is 0-49,999
# if it's greater than 371,000 seems data was updated
page_df <- tibble(a = 50000,
       b = seq(0, 400000, by = 49999)) 

payroll_list <- map2(
  page_df$a, page_df$b,
  ~ limit_path(.x, .y)) %>%
    map(get_parse)

# Drop rows for redundancy and in order to combine data frames
payroll_df <- payroll_list %>%
  map(~ select(., -mou, -payroll_department, - fms_department, -job_class)) %>%
  bind_rows() %>%
  distinct()
```

```{r}
foo <- payroll_list %>%
  map(~ select(., -mou, -payroll_department, - fms_department, -job_class))

# How do I write something like this
foo[map(foo, nrow) > 0] %>%
  bind_rows()

nrow(foo[[1]]) > 0

a <- foo[[1]]
a[nrow(a) > 0]

```



```{r data cleaning}
glimpse(payroll_df)

# Where is there missing data?
payroll_df %>%
  map_df(~ sum(is.na(.))) %>%
  select(where(~ . != 0)) %>%
  as.list()

payroll_df <- payroll_df %>%
  select(-ends_with("over_base_pay"))

payroll_df %>%
  count(mou_title, sort = TRUE) %>%
  print(n = 187)

```


## Testing to see how to narrow down categories
```{r tests}
payroll_df %>%
  select(mou_title) %>%
  mutate(mou_category = case_when(
    grepl('police', mou_title, ignore.case = TRUE) ~ 'Police',
    grepl('clerical', mou_title, ignore.case = TRUE) ~ 'Clerical',
    grepl('oper', mou_title, ignore.case = TRUE) ~ 'Operations',
    TRUE ~ as.character(mou_title)
    )
  )


payroll_df %>%
  select(mou_title) %>%
  mutate(mou_category = case_when(
    grepl('police', mou_title, ignore.case = TRUE) ~ 'Police',
    grepl('clerical', mou_title, ignore.case = TRUE) ~ 'Clerical',
    grepl('oper', mou_title, ignore.case = TRUE) ~ 'Operations',
    grepl('fire', mou_title, ignore.case = TRUE) ~ 'Fire',
    grepl('non-rep', mou_title, ignore.case = TRUE) ~ 'Non-Represented',
    grepl('admin', mou_title, ignore.case = TRUE) ~ 'Admin',
    grepl('safe', mou_title, ignore.case = TRUE) ~ 'Safety',
    grepl('rec', mou_title, ignore.case = TRUE) ~ 'Recreation',
    TRUE ~ as.character(mou_title))
  ) %>%
  count(mou_category, sort = TRUE)
# I should really check all of these for matches that are inaccurate


# checking if this tidyverse version works
# this version does not
payroll_df %>%
  select(mou_title) %>%
  mutate(mou_category = case_when(
    str_detect('police', mou_title) ~ 'Police',
    str_detect('clerical', mou_title) ~ 'Clerical',
    TRUE ~ is.character(mou_title)))

```


## Palette tests
```{r palette tests}
c25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown"
)

benes_palette_alt <- setNames(c(c25[1:4], c25[13]),
                              nm = c("Police", "Fire", "City", "DWP", "NA"))
```


## Visualize
```{r visualize}
benes_palette <- setNames(hue_pal()(4),
                          nm = c("Fire", "City", "Police", "DWP"))

show_col(hue_pal()(4))

# Basic histogram of total pay
ggplot(payroll_df) +
  geom_histogram(aes(x = total_payments, fill = benefits_plan),
                 bins = 20, alpha = 0.6) +
  scale_x_continuous(limits = c(0, 400000), labels = comma) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = benes_palette_alt) +
  labs(title = "Total Payments of LA City Employees, 2013-2018") +
  theme_gray()
```


For reference, looks like I'm cutting out these individuals
```{r removed values}
payroll_df %>%
  filter(total_payments <= 0 | total_payments >= 400000)

# 108 here, but only 86 or 96 for removed values
```
May want to filter this so that more of the x axis is used.
20 bins or so is a good size, not a mess of bins, but really need 15+ to see the details. Black outline is good but not used in this case for visibility, skew to the right


## Payments and benefits at the highest and lowest levels
I am curious about the total pay at the highest incomes and the lowest incomes. What benefits plan are they on? What does the shape of the distribution look like? I'll look at high earners first.


```{r 150000 histogram}
# Remove black outlines for visibility
payroll_df %>%
  filter(total_payments > 150000) %>%
  ggplot() +
  geom_histogram(aes(x = total_payments, fill = benefits_plan),
                 bins = 30, alpha = 0.7) +
  scale_x_continuous(labels = comma)
```


Compared to all employees, the high earners (150,000+) appear to have fewer folks on the city or the police benefits plans. But on the rightmost side of the distribution (300,000+), I see tiny slivers of blue and red.


```{r 300000 histogram}
payroll_df %>%
  filter(total_payments > 300000) %>%
  ggplot() +
  geom_histogram(aes(x = total_payments, fill = benefits_plan),
                 bins = 10, alpha = 0.7) +
  scale_x_continuous(labels = comma)
```


Most of the *very* high earners appear to be LAFD or DWP employees. But the *highest* earners look like they're all city employees. 


To be continued...